---
title: "Assessment"
author: "Paddy Kafeero"
date: "2026-01-28"
output: word_document
---

```{r setup, include=FALSE}
source("C:/Users/PaddyK/Desktop/Random_r_Code/src/setup.R")
```

Study 1 - cohort study

[Statistical Methods]{.underline}

Data was provided on the sex, age at enrolment into the cohort, smoking status at enrolment, death within 10 years of enrolment, indication of inclusion in the unmatched case-control study, indication of inclusion in the matched case-control study, and the matched set identifier for the matched case-control study.

Descriptive analysis was done using tabulations for the means, standard deviations, median and IQ of age, overall and by the smoking status at enrolment. Similarly, counts and frequencies for sex at enrolment and death within 10 years were provided, overall and by the smoking status.

Data was analysed

a)  Preliminaries Carry out and appropriately report any preliminary descriptive analysis of the data that you judge necessary.

```{r}
#Importing the data
smodo2026 <- rio::import("C:/Users/PaddyK/Desktop/MSc Medical Statistics/Stat_Models_Discrete/Assignment/smodo2026.csv")
##Checking for duplicate IDS
anyDuplicated(smodo2026$id)

#cohort <- smodo2026 %>% dplyr::filter(study2==0 & study3==0) %>% dplyr::select(id, sex, age, smoke, d)
cohort <- smodo2026 %>% dplyr::select(id, sex, age, smoke, d)
cohort <- cohort %>% mutate(
  sex = factor(sex, levels = c(0,1), labels = c("female", "male")),
  smoke = factor(smoke, levels = c(0,1), labels = c("non-smoker", "current smoker")),
  d = factor(d, levels = c(0,1), labels = c("survived", "died")),
  
)

#Exploring the data
str(cohort)
skim(cohort, age)

cohort %>% tabyl(smoke)
cohort %>% tabyl(d)



##pairwise comparison: 
cohort %>% tabyl(sex, d) %>% adorn_totals("col") #Mortality in the sex levels


cohort %>% tabyl(smoke, d) %>% adorn_totals("col")
cohort %>% group_by(d) %>% skim() %>% summarise()
plot(cohort$age, factor(cohort$d))
cohort %>% ggplot(aes(x=factor(d), y=age)) + geom_boxplot()

##Table1
#summarytools::dfSummary(cohort)


tab1<-CreateTableOne(
  vars=c("age","sex","smoke","d"),
  data=cohort,
  strata = "smoke",
  factorVars=c("sex","smoke","d"),
  includeNA = FALSE,
  test = F,
  addOverall = T
)
print(tab1,showAllLevels = T)
cohort %>% group_by(smoke) %>% summarise(medage=median(age), quantile(age, probs=c(0.25, 0.75)))

###Graphs
#Age descriptions
#Box plot of age and smoking status
age_smokeplot <- cohort %>% ggplot(aes(y=age, x=factor(smoke), fill=factor(sex))) +
  geom_boxplot(color="black",show.legend = F) +
  labs(x="Smoking status",
       y="Age at enrolment (years)",
       title = "Boxplot for age at enrolment and smoking status")

age_smokeplot
```

| Characteristic | Level | Overall | Non-Smoker at enrolment | Smoker at enrolment |
|---------------|---------------|---------------|---------------|---------------|
| N(%) |  | 10000 (100.0) | 3922 (39.2) | 6078 (60.8) |
| Age at enrolment (years) | Mean(SD) | 40.97 (15.40) | 46.68 (15.19) | 37.29 (14.37) |
|  | Median(IQR) | 39.00 (28.00-54.00) | 47.00 (34-61) | 34.00 (25-47) |
| Sex | Female | 5288 (52.9) | 1942 (49.5) | 3346 (55.1) |
|  | Male | 4712 (47.1) | 1980 (50.5) | 2732 (44.9) |
| Death within 10 years of enrolment | Survived | 9190 (91.9) | 3582 (91.3) | 5608 (92.3) |
|  | Died | 810 (8.1) | 340 (8.7) | 470 (7.7) |

b)  Unadjusted model (3%)

    Fit a logistic regression model relating smoking status to mortality ignoring age and sex.

```{r}
cohort_mod1 <- glm(d~factor(smoke), family = binomial(link="logit"), data=cohort)
summary(cohort_mod1)
tidy(cohort_mod1, exp=T, conf.int = T)

```

The log odds of death as a smoker are 0.12449 times lower than non smokers. There is no evidence for this association.

(c) Adjusted model (8%) Fit a logistic regression model relating smoking status to mortality adjusting for age and sex. Here, you may assume that age has a linear effect on mortality on the logit scale.

```{r}

cohort_mod2 <- glm(d~smoke+age+sex, family = binomial(link="logit"), data=cohort)
summary(cohort_mod2)
tidy(cohort_mod2, exp=T)

##For confounding, we need to also check for association of the other cvariates on the exposure varaible smoke
summary(glm(smoke~age, family=binomial, data=cohort))
summary(glm(d~age, family=binomial, data=cohort))
summary(glm(smoke~sex, family=binomial, data=cohort))
summary(glm(d~sex, family=binomial, data=cohort))

##Run anova to see which of the models is good
anova(cohort_mod1, cohort_mod2)

```

On adjusting for age and sex, we notice very strong evidence (p-value \<0.0001) of the association of smoking and death within 10 years of enrolment, with smokers having almost twice the odds of death within 10 years as the non-smokers (OR: 1.96; 95% CI 1.67 - 2.31). This indicates the confounding effect of age and sex on the relationship of smoking and death.

The anova test shows us that the adjusted model fits the data better and hence can be used to model predictions of deaths within the cohort.

(d) Model including interactions (10%)

    Add two-way interaction terms to your model in c) to allow the effect of smoking on mortality to depend on age and sex.Present results relating to the effect of smoking from each of your models in one or more tables and interpret these. Use graphs to display how relevant predicted probabilities of death differ by age and sex for the models in c) and d).

```{r}
#cohort_mod3 <- glm(d~(smoke*age), family = binomial(link="logit"), data=cohort) ##does increase in risk of death with increasing age differ between smokers and non-smokers?
#summary(cohort_mod3)

cohort_mod3 <- glm(d~(smoke*age)+(smoke*sex), family = binomial(link="logit"), data=cohort) ##Is the association between smoking and mortality age dependent, and is the association between smoking and mortality sex dependent?
summary(cohort_mod3) ##The interaction p-value of 0.898 indicates that we have no interaction in this relationship.
tidy(cohort_mod3, exponentiate = T)

##Graphs
##Getting the conditional probabilities
cohort_pred1 <- expand.grid(sex = c("female", "male"), smoke = c("non-smoker", "current smoker"), age=cohort$age)
##MODEL 3
# Get predicted probabilities (type = "response" gives Pr(Y=1))
cohort_pred1$prob <- predict(cohort_mod3, newdata = cohort_pred1, type = "response")
View(cohort_pred1)

###PLOT
ggplot(cohort_pred1, aes(x = age, y = prob, colour = smoke)) +
  geom_line(size = 1) +
  facet_wrap(~ sex) +
  labs(
    x = "Age (years)",
    y = "Predicted probability of death",
    colour = "Smoking status",
    title = "Predicted probability of death by age,\nsmoking status and sex (with interactions)"
  ) +
  theme_bw()


###MODEL 2
cohort_pred1$prob2 <- predict(cohort_mod2, newdata = cohort_pred1, type = "response")
View(cohort_pred1)

###PLOT
ggplot(cohort_pred1, aes(x = age, y = prob2, colour = smoke)) +
  geom_line(size = 1) +
  facet_wrap(~ sex) +
  labs(
    x = "Age (years)",
    y = "Predicted probability of death",
    colour = "Smoking status",
    title = "Predicted probability of death by age,\nsmoking status and sex (without interactions)"
  ) +
  theme_bw()

# > ggplot(cohort_pred1, aes(x = age, y = prob, colour = sex)) +
# +     geom_line(aes(interaction(smoke, sex), linetype = smoke)) +
# +     labs(
# +         x = "Age (years)",
# +         y = "Predicted probability of death",
# +         colour = "Smoking status",
# +         title = "Predicted probability of death by age,\nsmoking status and sex (with interactions)"
# +     ) +
# +     theme_bw()
# > ggplot(cohort_pred1, aes(x = age, y = prob, colour = sex)) +
# +     geom_line(aes(group=interaction(smoke, sex), linetype = smoke)) +
# +     labs(
# +         x = "Age (years)",
# +         y = "Predicted probability of death",
# +         colour = "Smoking status",
# +         title = "Predicted probability of death by age,\nsmoking status and sex (with interactions)"
# +     ) +
# +     theme_bw()








##We can combine the plots and put them on the same grid
# Example: age grid and four probability vectors
plot(age, p_female_nonsmoker, type = "l", col = "black",
     xlab = "Age", ylab = "Predicted probability of death")

lines(age, p_female_smoker,  col = "red")
lines(age, p_male_nonsmoker, col = "blue",  lty = 2)
lines(age, p_male_smoker,    col = "darkred", lty = 2)

legend("topleft",
       legend = c("Female non‑smoker", "Female smoker",
                  "Male non‑smoker", "Male smoker"),
       col    = c("black", "red", "blue", "darkred"),
       lty    = c(1, 1, 2, 2))






# #plot them
# library(ggplot2)
# ggplot(cohort_pred1, aes(x = age, y = prob, colour = smoke)) +
# geom_line(linewidth = 1) +
# labs(
# x = "Age(years)",
# y = "Probability of death",
# colour = "Smoking status"
# ) +
# theme_minimal()
##model 2
#plot(cohort$age, cohort_mod2$fitted.values, xlab="Age(years)", ylab="Probability of death")

##model 3
# agePlotData <- data.frame(age=cohort$age,
# pihat1=cohort_mod2$fitted.values,
# pihat2=cohort_mod3$fitted.values)
# #now order by dose
# agePlotData <- agePlotData[order(agePlotData$age),]
# plot(agePlotData$age,
#      agePlotData$pihat1,
#      xlab="Age (years)",
#      ylab="Fitted probability of death")
# points(agePlotData$age,
#        agePlotData$pihat2,
#        col="red")
# legend("topleft",
#        legend = c("Model2", "Model3"),
#        col= c("black", "red"),
#        pch= c(1, 1),
#        lty= c(0, 0),
#        bty= "n")


```

Coefficients for the Interaction model term estimate std.error statistic p.value <chr> <dbl> <dbl> <dbl> <dbl> 1 (Intercept) 0.000750 0.357 -20.2 1.60e-90 2 smokecurrent smoker 1.44 0.419 0.865 3.87e- 1 3 age 1.09 0.00580 15.0 5.53e-51 4 sexmale 1.34 0.120 2.46 1.40e- 2 : The odds of death within males is higher than females by 34%. 5 smokecurrent smoker:age 1.00 0.00697 0.270 7.87e- 1 6 smokecurrent smoker:sexmale 1.45 0.160 2.32 2.06e- 2 : The excess risk of death within 10 years associated with being a smoker is higher within the males by 45%. While age is associated with death (which makes sense), the risk of death associated with smoking is not affected by the change in age of an individual (OR 1.00; p-value=0.787)

Table 2: Crude and adjusted results for effect of smoking on mortality

| Model | Odds Ratio estimate for Smoking (95% CI) | p-value |
|----|----|----|
| Unadjusted | 0.88 (0.76, 1.02) | 0.09 |
| Adjusted without interactions | 1.96 (1.67, 2.31) | \<0.001 |
| Adjusted with interactions | 1.44 (0.64, 3.32) | 0.387 |

Model 2: Adding age and sex to the model increased the odds ratio of smoking from 0.88 to 1.96.

(e) Proportion of preventable deaths (6%) Use your model with interactions to estimate what proportion of deaths would have been prevented had no one smoked. You are not required to come up with a 95% confidence interval. Report the estimate and comment on your findings

```{r}
#Using model cohort_mod3 to get proportion of preventable deaths given no one smoked. We are getting a marginal effect for smoke=non-smoker
#Either we use the avg_predictions command which gives us the estimates of Pr(Y=1|do(X=0)) and Pr(Y=1|do(X=1)).
avg_predictions(cohort_mod3,
variables = "smoke") 
##In this case we want: Pr(Y=1|do(X=0): This is proportion of deaths if everyone never smoked. Basically the proportional of preventable deaths if everyone had not smoked.




#OR manually by creating a dataset where all values of smoke are non-smoker and we predict an estimate based on the modell 3
cohort_pred2 <- subset(cohort, select = c(sex, smoke, age))
cohort_pred2$smoke <- 0
#Manually calcualate predicted probabilities for each individual
cohort_pred2$prednon = with(cohort_pred2, plogis(-7.195613+0.362930*smoke+0.087075*age+0.001880*smoke*age*smoke+0.294445*sex+0.369424*smoke*sex))

cohort_pred2$prednon = with(cohort_pred2, plogis(-7.195613+(0.362930*smoke)+(0.087075*age)+(0.001880*smoke*age)+(0.294445*sex)+(0.369424*smoke*sex)))
mean(cohort_pred2$prednon)


##From the two methods, we can say that the proportion of preventable deaths had no one smoked is approximately 6% (0.0613). This is basically consired the risk of death were everyone not to have smoked.

##We can use the values of the avg_predictions to get a marginal risk ratio for death had everyone smoked compared to no one smoking.



####
(sum(cohort$d)-sum(cohort_pred2$prednon))/sum(cohort$d)

```

The proportion of deaths that would have been prevented had no one smoked is 6% (Estimate: 0.0613).

Study 2 - unmatched case-control study a) Preliminaries

```{r}
#Keeping the data for the unmatched study.
unmatched <- smodo2026 %>% dplyr::filter(study2==1) %>% dplyr::select(id, sex, age, smoke, d, study2)
View(unmatched)
summarytools::dfSummary(unmatched)
tab_unmatched<-CreateTableOne(
  vars=c("sex","age","smoke","d","study2"),
  data=unmatched,
  factorVars=c("sex","smoke","d","study2"),
  includeNA = T,
  test = F,
  addOverall = T
)
print(tab_unmatched,showAllLevels = T)

#Plots
#Smoking status vs age
#Smoking status vs sex
table(unmatched$smoke, unmatched$sex) ##from gere we see that we have more smokers that are male, but we also have more non-smokers as male
#Smoking status vs d
table(unmatched$smoke, unmatched$d) ##We can see that we have more survivors as the smokers, and we can also see that we have more deaths in the non-smokers
or <- (500*340)/(310*470) #odds of smokers dieing are 16.7% higher than non-smokers (OR: 1.167; 95% CI)

##The odds ratio of a sm

```

b)  Models

```{r}
unmatched_mod1 <- update(cohort_mod2, data=unmatched)
summary(unmatched_mod1)
tidy(unmatched_mod1)

##We have almost identical estimates because the data for the case control study is drawn from the cohort sample.
##Looking at the the standard errors and the confidence intervals comparing the cohort and the unmatched case control, we see that we have smaller SEs and tighter CIs for the cohort because of the fact that we have more data/observations.

```

c)  Proportion of preventable deaths

```{r}

unmatched_pred_all <- subset(smodo2026, select = c(sex, smoke, age))
unmatched_pred_all$smoke <- 0
#Manually calcualate predicted probabilities for each individual
unmatched_pred_all$prednon = with(unmatched_pred_all, plogis(-4.859638+0.598740*smoke+0.087520*age+0.499927*sex))
mean(unmatched_pred_all$prednon)

##Probability of preventable deaths is 0.318 if no one smoked. Basically preventable deaths proportion is 3% of all the participants of the original dataset.

```

Study 3 - matched case-control study a) Preliminaries - Check for the number of sets

```{r}
#Keeping the data for the matched study.
matched <- smodo2026 %>% dplyr::filter(study3==1) %>% dplyr::select(id, sex, age, smoke, d, study3, match_set)
View(matched)
matched <- arrange(matched, match_set)


head(matched, 10)

# tab_matched<-CreateTableOne(
#   vars=c("sex","age","smoke","d","study3", "match_set"),
#   data=matched,
#   factorVars=c("sex","smoke","d","study3", "match_set"),
#   includeNA = T,
#   test = F,
#   addOverall = T
# )
# print(tab_matched,showAllLevels = T)


##Summarising the smoking status numbers for the cases and controls
table(matched$smoke[matched$d==1]) #We have 470 smokers who are cases; and 340 non-smokers who are cases
table(matched$smoke[matched$d==0]) ##We have 337 smokers who survived; and 473 non-smokers who survied
#From this summary we can see that more smokers died, and more non-smokers survied. Smokers are more in the deaths.

#
matched <- matched %>% mutate(
  sex = factor(sex, levels = c(0,1), labels = c("female", "male")),
  smoke = factor(smoke, levels = c(0,1), labels = c("non-smoker", "current smoker")),
  d = factor(d, levels = c(0,1), labels = c("survived", "died")),
  
)

#Looking at the age of the cases and the controls ##this isnt relevant since they are matched by age. However, with this we confirm that they are in the same range and hence no difference
ggplot(matched, aes(x = d, y = age)) +
  geom_boxplot() +
  labs(
    x = "Case status",
    y = "age"
)
t.test(age~d, data = matched) ##confirm no difference in age between cases and controls.


## Age and smoking
ggplot(matched, aes(x = smoke, y = age)) + ##from this plot we can see that the smokers were actually younger than the non-smokers
  geom_boxplot() +
  labs(
    x = "Smoking status",
    y = "age"
)
t.test(age~smoke, data = matched) ##we have evidence that there is a difference in age between smokers and non-smokers

#Histogram of Age Differences in Matched Pairs (for Study 3 only): Demonstrates matching effectiveness (mean difference 2.6 years, all ≤5 years, as per design). Useful for preliminaries to confirm no major mismatches on age/sex.
ggplot(diff_df, aes(x = age_diff)) + 
  geom_histogram(binwidth = 1) + 
  labs(x = "Absolute Age Difference in Matched Pairs (years)", y = "Frequency") + 
  theme_minimal()

```

We can see that we have two rows for each set. The first row of the data is for the case and the second row is for the control.

```{r}
#Treating the smoking as the dependet variable
matched_mod1 <- glm(smoke~d, data = matched)
summary(matched_mod1) 
tidy(matched_mod1, exponentiate = T) ##This approach is invalid but we have evidence that peple who who died had higher odds of being smokers that those who survived.

summary(glm(smoke~d+factor(match_set), data = matched))$coefficients[1:2,] ##adjusting for the matching variable does not change the estimate for the log odds of smoking betweeen the deaths and survivors. additionally, the SE in th new model has slightly increased after accounting for paired nature of the data.

##From the two models, we can see what we expect. We had higher possibility of smoking among the people that died within 10 years.
#But we can do some conditional logistic regression to make our data more suitable for this matched case! See below!

```

Conditional regression model!

```{r}
##First, we reshape the data to make it suitable for the analysis. Reshape on case status
matchedWide <- reshape(matched, v.names=c("smoke", "age"), timevar="d",
                       idvar="match_set",direction="wide")
table(matchedWide$smoke.died,matchedWide$smoke.survived) ###We tabulate and work with the discordant pairs to do the McNemar's test
binom.test(276,(143+276)) ##from our McNemar's test, we can say that with the p-value<0.001, the two population proportions are actually different (NUll is that the population proportions are the same)
#ODDS RATIO FROM Mantel Hanzel is 276/143: odds of death for a smoker: 1.93; indicating that smokers have almost twice the odds of death than non-smokers
#95% CI for this odds ratio is got from the CI of the binom.test: L <- (0.6111198/(1-0.6111198)); U <- (0.7040425/(1-0.7040425)) -- For the MH, we use the CI from the McNemar's test to generate the CI of the MH Odds Ratio Estimate
#1.93(95%CI: 1.57 - 2.37)
##From this, we can say that the odds of being a case among smokers is 1.93 times the odds of being a case among non-smokers conditional on sex and age, as this is a matched study. The odds ratio doesnt contain the 1, hence we can say that there is evidence at the 5% level.
#However, just like below, the mcNemar's test is looking for an association basically and it gives the conditional odds ratio where conditioning is on the set variable.
##


##The results we get below are a confirmation of what we got above!
##Smoking as only predictor: Now here we look at the conditional logistic regression. For conditional regression, we need to use the clogit function from the survival library. Also, not that we do not change the shape of the data
matched_clogit <- clogit(d~smoke+strata(match_set), data = matched) ##with the clogit, we have
summary(matched_clogit) #Being a smoker multiplies the odds of death by 1.93 (95% CI 1.577, 2.362) conditional on sex and age at enrolment. A likelihood ratio test provides very strong evidence (p<0.0001) of the association between smoking and mortality (conditional on sex and age at enrolment)

#Extending the model to adjust for age
matched_clogit1 <- clogit(d~factor(smoke)+age+strata(match_set), data = matched)
summary(matched_clogit1) ##Including age in the model considers the residual confounding that could come up because of the age range given (cases were matched to controls within a range of 5 years)
matched_clogit1 <- glm(d~factor(smoke)+age+sex, data = matched)
summary(matched_clogit1)
exp(coef(matched_clogit1))

```
