---
title: "Session15"
author: "Paddy Kafeero"
date: "2026-02-09"
output: html_document
---


```{r setup, include=FALSE}
#source("C:/Users/PaddyK/Desktop/Random_r_Code/src/setup.R")

```

Aims
• Tounderstand how to interpret coefficients from multinomial and ordinal logistic regres
sion models.
• To understand how to calculate estimated probabilities from multinomial and ordinal
logistic regression models.
• To understand how to compare the fits of multinomial and ordinal logistic regression
models
3. What is the best way to model age (categorical or continuous)?

```{r}
multi <- rio::import("C:/Users/PaddyK/Desktop/MSc Medical Statistics/Stat_Models_Discrete/Practicals/Session15/multinomial.csv")
View(multi)
str(multi)
```


Question 1
```{r}
multi$sbpcat <- cut(multi$sbp,
                    breaks = c(0, 120, 140, 160, 300),
                    labels = c("Normal", "Prehypertension", "High1", "High2"),
                    right = FALSE,
                    ordered = T) # make it ordered
prop.table(table(multi$sbpcat))


#Centering age
multi$cen_age <- multi$ageyrs- 50

```


Question 2
```{r}
library(nnet)
multimod1 <- multinom(sbpcat~sex+cen_age, data=multi)
summary(multimod1)

##1 unit increase of age conditon

```



Question 3
```{r}
##Model
#
##For a female aged 70 years: Whats the probability of SBP High 1?
#age=70, sex==1

exp(-1.5405251+(0.07866921*70))/(1+(exp(-1.5405251+(0.07866921*70))+exp( -2.3134874+( 0.10677553*70))))

#Probability is 23%
```


Question 4
```{r}
#Females
femaledata <- data.frame(cen_age=multi$cen_age, sex="Female", ageyrs=multi$ageyrs, sbpcat=multi$sbpcat)
femaledata$fitted <- predict(multimod1, newdata = femaledata )

fitted_df <- cbind(multi, fitted(multimod1))
head(fitted_df)


# Filter females
females <- fitted_df %>% 
  filter(sex == "female") %>% 
  select(cen_age, starts_with("prob"))

# Convert to long format
fem_long <- females %>%
  pivot_longer(cols = starts_with("prob"),
               names_to = "sbp_category",
               values_to = "probability")


# Plot
ggplot(fem_long, aes(x = cen_age, y = probability, colour = sbp_category)) +
  geom_line() +
  labs(title = "Predicted SBP Category Probabilities vs Age (Females)",
       x = "Centred age",
       y = "Predicted probability")


plot( femaledata$cen_age, femaledata$fitted)
ggplot(femaledata, aes(x = ageyrs, y = fitted)) +  geom_line(size = 1)

```
For the plots.
As you get older you are more linkely to get into prehyper and then leave.


Question 5
```{r}
ordmod1 <- polr(sbpcat~sex+cen_age, data=multi)
summary(ordmod1)
```
logit(P(Y≤k))=θk​−(0.27952⋅sexMale+0.05744⋅cen_age)


Question 6
```{r}
#Woman aged 70 (centered is 20) who is in SBP cat 3
exp( 1.8744-(0.05744*20)) - exp( -0.1949-(0.05744*20))

```



Question 7
```{r}

```


Question 8
```{r}

```


For the ordinal model.
Males have higher logodds 0.27 conditionall on age.
The older you are conditonal on seex the more likely you are to have SBP.









