---
title: "Session13"
author: "Paddy Kafeero"
date: "2026-02-02"
output: html_document
---

```{r setup, include=FALSE}
source("C:/Users/PaddyK/Desktop/Random_r_Code/src/setup.R")

```

Aims
1. Understand how to fit models to count data using the glm command.
2. Understand how to compare such models.
3. Understand the concept of overdispersion, and how to identify it.

Question 1
```{r}
medpar <- rio::import("C:/Users/PaddyK/Desktop/MSc Medical Statistics/Stat_Models_Discrete/Practicals/Session13/medpar.csv")
str(medpar)
View(medpar)
head(medpar, 10)

summary(medpar$los)

##95% CI for mean length of stay
summarytools::descr(medpar$los)
(L <- mean(medpar$los) - 1.96*(sd(medpar$los)/sqrt(1495))) #we can use the 1.96 for large samples, otherwise we would have used the t0.975,n-1 for the CI
(U <- mean(medpar$los) + 1.96*(sd(medpar$los)/sqrt(1495)))

cat("The estimate for mean leangth of stay is", mean(medpar$los), "with CI",L,"to",U )
```
Minimum value is 116 and minimum is 1


Question 2
```{r}
medpar_pois1 <- glm(los~1, family = poisson, data=medpar)
summary(medpar_pois1)
 exp(2.287896) ##gives us the mean value of the length of stay
exp(confint.default(medpar_pois1))

#Here we have tighter CIS as we assume the variable is Poisson distributed

##The deviance is greater than the Df, so we can say that the model is incorrectly specfed for the data.
```



Question 3
```{r}
library(sandwich)
vcovHC(medpar_pois1) ##sandwich variance of the intercept from abve
sqrt(diag(vcovHC(medpar_pois1))) ##This gives us the SE for our estimate above!

(U_sandw <- coef(medpar_pois1)[1] + 1.96* 0.02319038)
(L_sandw <- coef(medpar_pois1)[1] - 1.96* 0.02319038)
 exp(U_sandw)
 exp(L_sandw)
 
 
 ##Basically if the poisson was holding, the mean and variance of los would have been the same, and the it is much more h
```


Question 4
```{r}
medpar_pois2 <- glm(los~type2+type3, family = poisson, data=medpar)
summary(medpar_pois2)

exp(coef(medpar_pois2))
exp(confint(medpar_pois2))


```
type2       0.237344   0.020902   11.36   <2e-16 *** -- howmuch the log mean increases if you are in type 22 compared to type 1. Length of stay is 27% higher (1.267) in type two than in type 1 (95% CI is)
type3       0.725361   0.025903   28.00   <2e-16 ***


Question 5
```{r}
medpar_pois3 <- glm(los~type2+type3+factor(age), family = poisson, data=medpar)
summary(medpar_pois3)


##Likelihood ratio test
anova(medpar_pois1, medpar_pois2) ##string evidence that medpar_pois2 is better than medpar_pois1

logLik(medpar_pois3) #This is the saturated model
logLik(medpar_pois2) ##This is the current model



```


Question 6
```{r}
medpar_nb1 <- glm.nb(los~type2+type3, data=medpar)
summary(medpar_nb1)


###Factor age
medpar_nb2 <- glm.nb(los~type2+type3+factor(age), data=medpar)
summary(medpar_nb2)

##Comapre the negative binomial models
anova(medpar_nb2, medpar_nb1)

##p-value 0f 0.77 indicates. Different because with the poison model , we assumed reasonable model.
#Here we allow for possiility of overdispersion and now we see the model above is not rreally relevant.
```


Question 7
```{r}

```


Question 8
```{r}


```









