---
title: "Session4"
author: "Paddy Kafeero"
date: "2026-01-19"
output: html_document
---

The aims of this session are:
• To estimate conditional treatment effects for both binary and continuous outcomes.
• To estimate marginal treatment effects for both binary and continuous outcomes using
standardization.
• To interpret estimates of marginal and conditional treatment effects.


```{r}
source("C:/Users/PaddyK/Desktop/Random_r_Code/src/setup.R")
```


Import the kidney_example data
```{r}
kidney <- rio::import("C:/Users/PaddyK/Desktop/MSc Medical Statistics/Stat_Models_Discrete/Practicals/Session4/kidney_example.csv")

names(kidney)
str(kidney)
anyNA(kidney)
```

PART A: Binary Outcome
Explore the tables
```{r}
tabyl(kidney, y, x)

kidney %>% dplyr::group_by(factor(z)) %>% tabyl(y, x)
```

Question 1
```{r}
#i
log_y <- glm(y~factor(x)+factor(z), family = binomial(link="logit"), data=kidney)
summary(log_y)

#ii
#kidney$x <- factor(kidney$x)
#kidney$z <- factor(kidney$z)
log_xz <- glm(y~x+z+I(x*z), family = binomial(link="logit"), data=kidney)

summary(log_xz)

```
Question 2
```{r}
#i
kidney$fit <- fitted.values(log_xz)
View(kidney)


# Create a small grid of the scenarios we want to predict
nd <- expand.grid(x = c(0, 1), z = c(0, 1))

# Get predicted probabilities (type = "response" gives Pr(Y=1))
nd$prob <- predict(log_xz, newdata = nd, type = "response")
print(nd)
View(nd)


#ii: Conditional Risk Differences
#for For Z=0
RD_z0 <- nd$prob[nd$x==1 & nd$z==0] - nd$prob[nd$x==0 & nd$z==0]
RD_z1 <- nd$prob[nd$x==1 & nd$z==1] - nd$prob[nd$x==0 & nd$z==1]
#iii: Conditional risk ratio
RR_z0 <- nd$prob[nd$x==1 & nd$z==0] / nd$prob[nd$x==0 & nd$z==0]
RR_z1 <- nd$prob[nd$x==1 & nd$z==1] / nd$prob[nd$x==0 & nd$z==1]

```

Question 3
```{r}




```


Question 4
```{r}

#4i
avg_predictions(log_xz, variables = "x")
#Pr(Y=1|do(X=0))=0.169
#Pr(Y=1|do(X=1))=0.223

#4ii
avg_comparisons(log_xz, variables = "x")

```







Question 5
```{r}
pred_covs <- subset(kidney,select=c(x,z))
#set treatment to 0
pred_covs$x <- 0
#manually calculate predicted probabilities for each individual
#use plogis(x)=exp(x)/(1+exp(x))
pred_covs$pred0 = with(pred_covs,
plogis(-2.60269+0.7308873*x+1.607874*z-0.5245292*x*z))
#now set treatment to 1
pred_covs$x <- 1
pred_covs$pred1 = with(pred_covs,
plogis(-2.60269+0.7308873*x+1.607874*z-0.5245292*x*z))
#calculate the mean predicted outcome in each scenario; ##We get prob y when x=0 and prob y when x=1, then we average them and we see we get the same as above in the avr!!

mean(pred_covs$pred0)
mean(pred_covs$pred1)
#risk difference is
mean(pred_covs$pred1)-mean(pred_covs$pred0)



##We get prob y when x=0 and prob y when x=1

##How do we get the confidence interval for this?
library(boot)
quantile(boot_fit$t[,2], c(.025, .975))
boot::boot.ci(boot_fit, type = 'norm', index = 2)


```



PART B: CONTINUOUS OUTCOME


```{r}
kidney_cont <- rio::import("C:/Users/PaddyK/Desktop/MSc Medical Statistics/Stat_Models_Discrete/Practicals/Session4/kidney_example_continuousY.csv")

names(kidney_cont)
str(kidney_cont)
View(kidney_cont)
```



Question 6
```{r}

kidney_cont %>% 
  group_by(x) %>% 
  summarise(mean_y=mean(y))

kidney_cont %>% 
  group_by(z) %>% 
  summarise(mean_y=mean(y))

kidney_cont %>% skimr::skim(factor(kidney_cont$x), factor(kidney_cont$y))



```

Question 7
```{r}
kidney_cont$x <- factor(kidney_cont$x)
kidney_cont$z <- factor(kidney_cont$z)


lin_xz <- lm(y~x+z+(x*z), data=kidney_cont)
summary(lin_xz)

broom::tidy(lin_xz)
```


Question 8
```{r}

```

Question 9
```{r}

```


Question 10
```{r}

```


Question 11
```{r}

```





























