---
title: "Session6"
author: "Paddy Kafeero"
date: "2026-01-20"
output: html_document
---

```{r setup, include=FALSE}
source("C:/Users/PaddyK/Desktop/Random_r_Code/src/setup.R")

```



Part A = = = = %>% 
Part B
```{r}
cattaneo2 <- rio::import("C:/Users/PaddyK/Desktop/MSc Medical Statistics/Stat_Models_Discrete/Practicals/Session6/cattaneo2.dta")
str(cattaneo2)

cattaneo2 <- zap_labels(cattaneo2) ##zap_labels: Removes value labels, leaving unlabelled vectors as is. Use this if you want to simply drop all labels from a data frame.
cattaneo2$prevnewborndeath <- cattaneo2$deadkids
```

Question 1
Explore the dataset
```{r}
View(cattaneo2)
statar::sum_up(cattaneo2)
broom::tidy(cattaneo2)
summary(cattaneo2)
library(skimr)
skim(cattaneo2)  


##Checking levels
levels(cattaneo2$medu) ##number of years of education!

##b combinining variables: combined mrace and mhisp to get one race variable
table(cattaneo2$mhisp)
table(cattaneo2$mrace)
cattaneo2 <- cattaneo2 %>% mutate(race = case_when(
  mhisp == 1 ~ "Hispanic",
  mrace == 1 ~ "White",
  mhisp==0 & mrace==0 ~ "Other",
))

table(cattaneo2$race)
anyNA(cattaneo2$race
      )
```


Outcome: lbweight : 1 if low birthweight
Main Exposure: mbsmoke
Potential confounders: alcohol, race = (mrace+mhisp), mage, nprenatal, 

Question 2
```{r}
table(cattaneo2$mbsmoke, cattaneo2$lbweight)
cattaneo2$mbsmoke <- factor(cattaneo2$mbsmoke)

mod1 <- glm(lbweight~mbsmoke, family = binomial(link = "logit") , data = cattaneo2)
summary(mod1)
anova(mod1)
broom::tidy(mod1, exponentiate = T, conf.int = T)

```
The anova on model one confirms that there is truly an association between maternal smoking status and low birthweight.
However, we see that the odds of being low birthweight are two times higher (OR: 2.399) higher within the smoking mothers.

Question 3
```{r}
mod2 <- glm(lbweight~mbsmoke + mmarried + race + mage + nprenatal + medu + alcohol, family = binomial(link = "logit") , data = cattaneo2)
summary(mod2)
broom::tidy(mod2, exponentiate = T, conf.int = T)


predict(mod2,newdata=data.frame(mbsmoke=c(0,0,1,1),z=c(0,1,0,1)),type="response")


##Marginal mean differences
library(marginaleffects)
comparisons(
mod2,
newdata = "mean",
variables = list(mbsmoke= c(0,1)), 
df="residual"
)


library(marginaleffects)
comparisons(
mod2,
newdata = "mean",
variables = "mbsmoke", 
df="residual",
transform = "exp"
)

avg_comparisons(mod2, comparison = "difference") # average (marginal) estimates.

```
Estimated marginal mean difference: odds 1.028807
The smokers have 3% (2.8%) higher chance of low birth weight babies than the non smokers.


Question 4
Models with lower AIC (Akaike Information Criterion) are preferred to those with higher values.
Here we do forward selection by AIC to select confounders to adjust for.
```{r}
#model with just mbsmoke
mbsmokeMod <- glm(lbweight~mbsmoke, data=cattaneo2, family="binomial")

step(mbsmokeMod, 
     direction = "forward", 
     scope = list(lower = ~mbsmoke, upper = ~mbsmoke + mmarried + mhisp + fhisp + foreign + 
                    alcohol + prevnewborndeath + mage + medu + fage + fedu + monthslb + 
                    order + mrace + frace + birthmonth))


mod_forward <- glm(formula = lbweight ~ mbsmoke + mrace + mmarried, family = "binomial", 
    data = cattaneo2)
avg_comparisons(mod_forward, comparison = "difference")

```


Question 5
```{r}
install.packages("chest")
library(chest)

chest_glm(crude = "lbweight ~ mbsmoke",
                 xlist = c("mmarried", "mhisp", "fhisp", "foreign", 
                    "alcohol", "prevnewborndeath", "mage", "medu", "fage", "fedu",
                      "monthslb", "order", "mrace", "frace", "birthmonth"),
                 data = cattaneo2) ##thos one is using the p-change in estimates to come up to the most relevant values that change the estimate the most
summary(mod4)

```


Question 6
```{r}
mbsmokeMod2 <- glm(lbweight~mbsmoke, data=cattaneo2, family="binomial")
step(mbsmokeMod2, 
     direction = "both", 
     scope = list(lower = ~mbsmoke, upper = ~mbsmoke + mmarried + mhisp + fhisp + foreign + 
                    alcohol + prevnewborndeath + mage + medu + fage + fedu + monthslb + 
                    order + mrace + frace + birthmonth))


mod_both <- glm(formula = lbweight ~ mbsmoke + mrace + mmarried, family = "binomial", 
    data = cattaneo2)
summary(mod_both)
broom::tidy(mod_both, exponentiate = T, conf.int = T)

###GIVE MARGINAL RISK DISFFERENCE
avg_comparisons(mod_both, comparison = "difference", vcov = "HC1")
#Marginal risk difference is 0.048 (CI)

```


Question 7
```{r}
##Question 7 usess the LASSO method!


```







