---
title: "Session9"
author: "Paddy Kafeero"
date: "2026-01-26"
output: html_document
---


```{r setup, include=FALSE}
source("C:/Users/PaddyK/Desktop/Random_r_Code/src/setup.R")

```

Aims
• Learn how to fit a conditional logistic regression model to matched data.
• Learn how to fit a standard logistic regression model to matched data by including the matching variables.
• Understand the different assumptions underpinning the two models described above.


Part A: Vitamin E dataset

```{r}
vit <- rio::import("C:/Users/PaddyK/Desktop/MSc Medical Statistics/Stat_Models_Discrete/Practicals/Session8/vit_E.csv")
str(vit)

head(vit, 10)

vit$vitEHigh <- ifelse(vit$vitE>12, 1, 0)

```


Question 1
```{r}
mod1 <- glm(case~vitEHigh+factor(set), family = binomial, data = vit)
summary(mod1)

tidy(mod1, exponentiate = T)
```
We can see that the odds ratio of 0.57 obtained here is not the same as the 0.76 that we got earlier.
From the notes we see that this is the incoreect approach to getting this value

Question 2
```{r}
##Fitting a conditional odds ratio: We use the survival package.
mod2 <- clogit(case~vitEHigh + strata(set), data = vit)
summary(mod2)


```
Odds of being a case are lower when you have high vitamin than with low vitamin E

Conditioning on high vitamin C here, we see that the odds ratio is similar to the one from the previous lecture, showing the relevance of the  indicated format.
Also, worth noting is when we compare this to the incorrect model, squaring the vitEHigh estimate will give us the value of the incorrect model. The factor varianle alters the MLE for vitEHigh

With this we see that for our binary dependent variables in matched data, it is important to use the conditional logistic regression rather than the normal one!

Question 3
```{r}
mod3 <- clogit(case~vitE + strata(set), data = vit)
summary(mod3)
##this ana
```
Odds of being a case decrease by 1% for a unit increase in vitE. 

Question 4
```{r}
##Suppose we want to explore effect of dichotomised vitamin E on conditional odds
mod4 <- clogit(case~vitEHigh*factor(time) + strata(set), data = vit)
mod4 <- clogit(case~vitEHigh:factor(time), data = vit)

summary(mod4)

##doing an anova between the model before and this one to check whether there is evidence of effect of time across the 3 llevels?


```

time: a unit increase in time, there is a reduced 1% chance of being a cance case. Indicating that the longer the observation time, the lower the chances of being a cancer case.

We want to look at whether exposure effect differs by matching factors.



PART B. INFERTILITY DATASET
Research Question: DO PREVIOUS ABORTIONS AFFECT THE RISK OF INFERTILITY


Question 5
```{r}
infertility <- rio::import("C:/Users/PaddyK/Desktop/MSc Medical Statistics/Stat_Models_Discrete/Practicals/Session9/infertility.csv")
str(infertility)

head(infertility, 10)

##We have one person per row
table(infertility$case)
#83 cases
#
table(infertility$stratum)
##all the sets are the same size!!!

##Also explore the variables to see if there is really perfect match on number of controls vs cases

```


Question 6
```{r}
mod5 <- clogit(case~factor(spontaneous) + strata(stratum), data = infertility)
summary(mod5)

##if we had put spomt as not factor, we would be assuming a linear effecr of spontaneous on case!

```
factor(spontaneous)1     3.112: For having 1 abortion, the odds of being infertile are 3.122 times those with no abortions
factor(spontaneous)2     7.993: For having 2 or more abortions, the odds of being infertile are 7.993 times those with no abortions

From these, we can say that the chances of infertility are conditional on the number of abortions!

Question 7
We

```{r}
#a
mod6 <- glm(case~age+parity+education, family = binomial, data = infertility)
tidy(mod6, exp=T)

#b
mod7 <- glm(case~factor(spontaneous), family = binomial, data = infertility)

tidy(mod7, exp=T) ##in this model we are not accounting for the matching factors
#In the part c, we account for the matching factors


#c
mod8 <- glm(case~factor(spontaneous)+age+parity+education, family = binomial, data = infertility)
tidy(mod8, exp=T)

##IN this mdoel you will even see that the standard error is lower than the original clogit.


```


First model: Checking association of matching variables with outcome. We see no association, indicating independence of the matching variables from the outcome.
Second model:
Third mode: COnditional on exposure there matching variables are no longer independent of the outcome hence showing yhe importance of adjusting for them. Compared to the first model, the p-values in this one show the change when we include the exposure variable in the model.








