---
title: "Session14"
author: "Paddy Kafeero"
date: "2026-02-09"
output: html_document
---

```{r setup, include=FALSE}
#source("C:/Users/PaddyK/Desktop/Random_r_Code/src/setup.R")

```

Aims
The aim of this session is to address the following questions about these data:
1. Is there a difference between deaths rates in the two factories?
2. Is death rate associated with age?
3. What is the best way to model age (categorical or continuous)?

```{r}
rubber <- rio::import("C:/Users/PaddyK/Desktop/MSc Medical Statistics/Stat_Models_Discrete/Practicals/Session14/rubber.csv")
View(rubber)
```


Question 1
```{r}
#Total number of deaths
tdeaths <- sum(rubber$deaths)

#Total number of person years
tpyrs <- sum(rubber$pyrs)

#Overall death rate
(drate <- tdeaths/tpyrs)
cat("Overall rate of death is", round(drate*1000, 2),"deaths per 1000 person-years")

#Log death rate
(ldrate <- log(drate))
cat("Log death rate is", ldrate)

```


Question 2
```{r}
##Death rate in each factory
sumtab <- rubber %>% group_by(factory) %>% summarise(tdeaths=sum(deaths), tpyrs=sum(pyrs), deathrate=sum(deaths)/sum(pyrs))

f1drate <- 0.007366483
f2drate <- 0.009194442

lf1drate <- log(f1drate)
lf2drate <- log(f2drate)


```


Question 3
```{r}
##Death rate variable per row
library(dplyr)
rubber <- rubber %>% group_by(factory) %>% mutate(log_death_rate = log(deaths/pyrs) )
head(rubber)

##Plot log death rates against age group
library(ggplot2)
rubber %>% ggplot(aes(agegrp, log_death_rate)) +
  geom_point(size = 3) +
  geom_text(aes(label = factory), position = position_nudge(x = 0.1)) +
  labs(x = "Age group", y = "Log death rate")

```
This plot suggests that death rate decreases with increasing age group. It also suggests that factory 1 has higher death rate than factory 2.

Question 4

Its a Poisson distribution.

$log(deathrate_{f0, Ag50-59})=log(pyrs) + \beta_{0}+\beta_{1}factory_i+\beta_{2}Age_{60-69}+\beta_{3}Age_{70-79}+\beta_{3}Age_{80-89}$
$log(deathrate_{Ag50-59})=log(pyrs) + \beta_{0}+\beta_{1}Age_{60-69}+\beta_{2}Age_{70-79}+\beta_{3}Age_{80-89}$

Question 5
```{r}
#-----a

#Model without age
poissmod <- glm(deaths~offset(log(pyrs)), family = poisson, data=rubber)
summary(poissmod) ##oly the model with death rate

#
poissmod1 <- glm(deaths~factor(agegrp), family = poisson, data=rubber, offset = log(pyrs))
summary(poissmod1)
tbl_regression(poissmod1,
    exponentiate = TRUE, conf.int = T
  )

#compare the models
anova(poissmod1, poissmod)
#test statistic here is differentce in the log likelihooods * 2

#for agegroup 1: 
#Very strong evidence of association of age and death rate, for all the age groups.


#-----b
#---i
rr_6069_5059 <- exp(coef(poissmod1)[2])
drate_6069 <- coef(poissmod1)[1]+coef(poissmod1)[2] ##Death rate for the agegroup 60-69
drate_5059 <- coef(poissmod1)[1]
exp(drate_6069)/exp(drate_5059) ##Confidence interval: 2.82 - 9.04
Urr_6069_5059 <- exp(coef(poissmod1)[2]+1.96*(0.2950))
Lrr_6069_5059 <- exp(coef(poissmod1)[2]-1.96*(0.2950))


exp((coef(poissmod1)[1]+coef(poissmod1)[2])-coef(poissmod1)[1])
confint(poissmod1)

#---ii
drate_7079 <- coef(poissmod1)[1]+coef(poissmod1)[3]
drate_8089 <- coef(poissmod1)[1]+coef(poissmod1)[4]
exp(drate_8089)/exp(drate_7079) ##Rate ratio for 80-89 vs 70-79


###SE for ii
vcov(poissmod1)

SE <- sqrt()


##Also the marginal effects package can be used to get that: Use the hypothesis function


```


Question 6
```{r}
poissmod2 <- glm(deaths~factor(agegrp)+factor(factory), family = poisson, data=rubber, offset = log(pyrs))
summary(poissmod2)
tbl_regression(poissmod2)
####
anova(poissmod2, poissmod1)



##Rate ratio for the two factories
exp(coef(poissmod2)[5])
exp(coef(poissmod2)[5]-(1.96*0.1590)); exp(coef(poissmod2)[5]+(1.96*0.1590)) ##95% CI

```
There is no evidence of difference in rates for the two factories (p-value=0.223). Thi sshouldnt suprise us given that he residual deviance in the model fo age alone showed that there is no evidence of refsing the model. it was alomist perfect





Question 7
```{r}
#Interaction model
poissmod3 <- glm(deaths~factor(agegrp)*factor(factory), family = poisson, data=rubber, offset = log(pyrs))
summary(poissmod3)
tbl_regression(poissmod3, exp=T)
#a This is the SATURATED model: The deviance is extremely big and the Df =0. This is the saturated model, it uses up all the DFs
#b No evidence of interaction



```


Question 8
```{r}

```


Question 9
```{r}

```











