---
title: "Session7"
author: "Paddy Kafeero"
date: "2026-01-20"
output: html_document
---


```{r setup, include=FALSE}
source("C:/Users/PaddyK/Desktop/Random_r_Code/src/setup.R")

```

Aims
We will examine the association between oesophageal cancer as the outcome and a dichotomous smoking status variable (smoker or non-smoker) as an exposure, and attempt to answer the following questions:
1. Does smoking increase the risk of oesophageal cancer?
2. Does alcohol intake or age confound this relationship?
3. What is the best way to adjust for these potential confounders: as categorical or continuous variables?
We will also swap the exposure and outcome variables to explore what happens to the estimate of the odds ratio (and its confidence interval)



Analysis
Question 1

```{r}
oeso1 <- rio::import("C:/Users/PaddyK/Desktop/MSc Medical Statistics/Stat_Models_Discrete/Practicals/Session7/oeso1.csv")
str(oeso1)

oeso1$tobacco_2 <- ifelse(oeso1$tobacco_group==0, 0, 1)
##Associations between binary tobacco status and alcohol group
oeso1 %>% tabyl(tobacco_2, alcohol_grp) ###check using a heat map!!!!
mod1 <- glm(tobacco_2~factor(alcohol_grp), family=binomial, data = oeso1)
tidy(mod1, exponentiate = T, conf.int = T)
##Associations between binary tobacco status and age
oeso1 %>% tabyl(tobacco_2, age_group)
mod2 <- glm(tobacco_2~age_group, family=binomial, data = oeso1)
tidy(mod2, exponentiate = T, conf.int = T)
##Associations between alcohol and age
oeso1 %>% tabyl(alcohol_grp, age_group)
oeso1$alcohol_grp<-factor(oeso1$alcohol_grp)
oeso1$age_group<-factor(oeso1$age_group)
mod3 <- glm(alcohol_grp~age_group, family="binomial", data = oeso1)
tidy(mod3, exponentiate = T, conf.int = T)

```


Question 2
```{r}
#a
tob_case_tb<-oeso1 %>% tabyl(case, tobacco_2) %>% adorn_percentages() %>% adorn_ns()
View(tob_case_tb)

#ODD ratio
(or_cc<-(255*191)/(9*521))

var<-(1/255)+(1/191)+(1/521)+(1/9)

Ci_L<-or_cc - (1.96*sqrt(var))
Ci_U<-or_cc + (1.96*sqrt(var))

CI <- c(Ci_L, Ci_U)
cat("The estimate is", or_cc, "with CI: ", Ci_L, "to", Ci_U)

#b


```


Question 3
```{r}
#a
mod4<-glm(case~tobacco_2, family = binomial, data=oeso1)
tidy(mod4, exponentiate = T, conf.int = T)

#b
mod5<-glm(tobacco_2~case, family = binomial, data=oeso1)
tidy(mod5, exponentiate = T, conf.int = T)
```
For the case control study we see that the odds ratios are the same regardless of the order of the main exposure and the outcome.


Question 4
```{r}
#a
mantelhaen.test(x=oeso1$tobacco_2, y=oeso1$case, z = oeso1$alcohol_grp)

##OR: cREATE A TABLE AND CALCULATE MANUALLY
oeso1$cases<-factor(oeso1$cases)
tab <- xtabs(~ oeso1$case + oeso1$tobacco_2 + oeso1$alcohol_grp, data = oeso1)
num<-(((148*29)/(237+29+148))+((80*72)/(201+72+80+3))+((22*46)/(66+46+22+5))+((5*44)/(17+44+5+1)))
denom<-(((0*237)/(237+29+148))+((3*201)/(201+72+80+3))+((5*66)/(66+46+22+5))+((1*17)/(17+44+5+1)))
or_mh<-num/denom

#b
mantelhaen.test(x=oeso1$tobacco_2, y=oeso1$case, z = oeso1$alcohol_grp)

```

Question 5
```{r}
#a
mod6 <- glm(case~tobacco_2+alcohol_grp, family=binomial, data=oeso1)
tidy(mod6, exponentiate = T, conf.int = T)

#b
mod7 <- glm(tobacco_2~case+alcohol_grp, family=binomial, data=oeso1)
tidy(mod7, exponentiate = T, conf.int = T)

```
Value slightly different because in the glm its the mle we get while in the mh we dont get the mle. But they are both valid estimates.

Question 6
```{r}
mod8 <- glm(case~tobacco_2+alcohol_grp+age_group, family=binomial, data=oeso1)
tidy(mod8, exponentiate = T, conf.int = T)


```


Question 7
```{r}

```



Question 8
```{r}

```









